---
title: "Introduction to roboplotr"
output: rmarkdown::html_vignette
description: >
  Start here for an overview of roboplotr. This will cover the basic idea
  why the package exists, and shows you basic usage or the core function,
  roboplot.
vignette: >
  %\VignetteIndexEntry{Introduction to roboplotr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opt_chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(roboplotr)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Basics

Plotly is especially challenging to use when the resulting plots are dynamically
sized, as elements external to the plot area, like titles, captions, legends and 
logos will not be located where they are expected to be after a resize. The 
purpose of the roboplotr package is to reduce the workload of making dynamic 
plotly plots, with many layout elements handled with parameters to the core 
roboplotr function, `roboplot()`. Some more complicated things are handled with 
helper functions. Roboplotr has internal scripts to handle plot relayouts on a 
resize, including dynamic title wraps and y-axis font resizing.

This vignette will show an example how to use roboplotr package to create 
an html report with several interactive plots in one go with minimum hassle for
the plots themselves, and we will also capture all the created plots as separate 
files in png format. For the purposes of this report we will use set widths for
all the plots, but roboplot truly shines when you need your plots to fit 
different device widths, in which case you would not give set widths to the
plots.

Every function parameter we use here (and the ones we don't) have an example of
use in the function documentation.

## Setting the global options

First off, we use `set_roboplot_options()` to set some global options to define 
some visual and other specifications. We create the directory "plots" in tempdir()
to store any files created in.

```{r roboplot_options}

dir.create(file.path(tempdir(), "plots"))

set_roboplot_options(
  artefacts = set_artefacts(
    filepath = file.path(tempdir(), "plots"),
    auto = T,
    artefacts = "png"
    ),
  caption_template = "Source: {text}.",
  height = 400,
  imgdl_wide = set_imgdl_layout(
    width = 650,
    height = 400,
    suffix = "",
    format = "png"
  ),
  imgdl_narrow = set_imgdl_layout(
    width = 550,
    height = 440,
    suffix = "",
    format = "png"
  ),
  modebar = c(
    "home",
    "closest",
    "compare",
    "zoom",
    "img_n",
    "img_w",
    "data_dl"
  ),
  logo_file = system.file("images", "robonomist.png", package = "roboplotr"),
  trace_colors = grDevices::palette.colors(6, "Set2"),
  width = 650,
  xaxis_ceiling = "guess",
  verbose = "Warning"
)
```

Above we defined the logo used in the plots, defined how captions are 
constructed (avoiding repetedly writing the same information). Modebar
elements were also defined, with further specifications  for image downloads for
the external png files. We also defined the colors used by all plot traces in 
the order we want them to be used, as well as the plot dimensions, and the 
x-axis ceiling handling. 

When `xaxis_ceiling` is provided, any scatter plots with date format are given
some breathing room to the upper end of the range by rounding the last date unit 
to the next unit of the denomination given, ie. if the last date value is 
`"2022-10-07"`, the x-axis upper bound will be `"2022-11-01"`.

Captions, height, width, xaxis_ceiling, trace colors and artefacts can be 
controlled by-plot, too, but giving them here avoids writing repeatedly. As 
`roboplot()` is written with dynamic plot widhts in mind, it's usually unnecessary
to define it, but for demonstration purposes we provided a width, as we want our
rendered plots to be the same width as the corresponding png files we are going
to create.

The specifications for image downloads are defined with `imgdl_wide` and 
`imgdl_narrow`. These parameters define the desired image dimensions for static
image download buttons in the modebar,  and provide a suffix for the created files, 
and define the file format. We set these to match our intended programmatically
created .png files.

We also want to automate the artefact file creation. This can be controlled 
by-plot with the parameter `artefacts` of `roboplot()`, but we'll use 
`set_artefacts()` here to set some global options. We set `auto` to true
(to make `roboplot()` create the files automatically when creating a plot), and 
give `filepath` where the files will be located. As all our plots have height 
and width set, these will be used in the resulting static plots, too, unless
overridden by `artefacts`.

We also set `verbose` to "Warning" that quiets most of the messages roboplotr
gives.

Next we call some libraries we need for the data we are going to plot, and 
create our first plot.

## Basic use

```{r gdp}
library(tidyverse)
library(robonomistClient)
library(lubridate)

d <- data_get("oecd/EO") |>
  filter(
    Country %in% c("Finland", "Euro area (17 countries)", "United States", 
                   "United Kingdom", "Japan"),
    Frequency == "Quarterly",
    Variable == "Gross domestic product, volume, market prices",
    time >= "2017-01-01"
  ) |>
  collect() |>
  with_groups(
    Country, mutate,
    value = 100 * value / mean(value[year(time) == 2019])
  ) |>
  mutate(Country = recode(Country, "Euro area (17 countries)" = "Euro area"))

attr(d, "frequency") <- "Quarterly"

roboplot(d,
    Country,
    title = "Real GDP forecasts",
    subtitle = "Index, 2019=100",
    caption = "OECD"
  )

```


At the minimum, `roboplot()` needs some data to be plotted, a title, and a 
caption. Subtitle is optional, but most likely needed - it is the best place to
show the y-axis unit. Caption is parsed from the given string and elements 
defined in `set_roboplot_options()`.

Data transformation also stripped the data of the frequency attribute, which 
`roboplot()` uses to parse date formats for several purposes, so we inserted that
back. We could have skipped this, too, leaving it up to `roboplot()` to guess the 
frequency.

## Caption control

Next we will create another plot, with a caption that does not respect the
globally set format.

```{r inflation}
inflation_eu <-
  data_get("eurostat/prc_hicp_manr") |>
  filter(
    unit == "Annual rate of change",
    coicop == "All-items HICP",
    geo %in% c("Germany (until 1990 former territory of the FRG)",
      "Euro area - 19 countries  (from 2015)","Finland","Sweden"),
    time >= "2015-01-01"
  ) |>
  mutate(
    geo = recode(geo,
      "Euro area - 19 countries  (from 2015)" = "Euro area",
      "Germany (until 1990 former territory of the FRG)" = "Germany"
    )
  ) |>
  select(geo, time, value)

inflation_us <-
  data_get("fred/CPIAUCSL", units = "pc1") |>
  drop_na() |>
  filter(time >= "2015-01-01") |>
  transmute(geo = "USA", time, value)

d <- bind_rows(inflation_us, inflation_eu)

d |> roboplot(geo, "Consumer price inflation","Annual change of HICP, %",
                  caption = set_caption(template = "Sources: Eurostat & US BLS."))
```

The previous examples show how plotting with roboplotr is a lightweight task. As
the source notation is a bit different from the previous plot and we have
predefined caption formatting previously with `set_roboplot_options()`, we must
use `set_caption()` to override the prefix instead of just giving a single string.

## Horizontal bar plots and x-axis string width, and omitted legends

Next we make a horizontal plot on producer prices, where we need to be be a bit
more verbose with plot axes.

```{r producer_prices}
d <-
  data_get(
    "StatFin/thi/statfin_thi_pxt_118g.px", lang = "en", tidy_time = TRUE) |>
  filter(str_detect(
    `Products by activity (CPA 2015, MIG)`,
    "^(01|02|03|07|08|10|16|17|19|20|21|22|23|24|25|26|27|28|29|30|35|36|42) "),
    `Index series` %in% c("Producer price index for manufactured products"),
    Information %in% c("Annual change, %")
  ) |>
  filter(time == max(time)) |>
  rename(CPA = `Products by activity (CPA 2015, MIG)`) |>
  drop_na(value) |>
  mutate(CPA = fct_reorder(str_replace(CPA, "[0-9]{2} ", ""), value))

d_title <- str_c("Producer prices in Finland ",format(unique(d$time), "%m/%Y"))

d |> roboplot(
  title = d_title, 
  subtitle = "%, annual change", 
  caption = "Statistics Finland",
  plot_type = "bar",
  plot_mode = "horizontal",
  plot_axes = set_axes(y = "CPA"),
  legend_maxwidth = 45
  )
```

As we wanted to present a horizontal bar plot, we used the parameters 
`plot_type` and `plot_mode` from `roboplot()` to define the plot as such.

Before plotting, we reordered the column that would be y-axis as factor based
on value, as `roboplot()` only does this independently for the column that defines 
`color`.

We also used `plot_axes` parameter of `roboplot()` and the function 
`set_axes()` parameter `y` to set the column "CPA" from the data to be the
yaxis of the plot. set_axes() can take further parameters (described in
detail in function documentation), but for a horizontal bar plot where you have 
a numeric column named `"value"`, `set_axes()` can handle the ticktypes etc. for the
plot by just providing y-axis data column name as parameter `y`. For most cases
this is enough.

As we *didn't* want to color-code the bars, we skipped providing
`color` for `roboplot()`. As there is now only a single trace color, `roboplot()` 
omitted the legend. We could also have provided another column for `color` that
only has a single observation, and the result would have been the same. We could
have also included the legend with setting `legend_position` to `"bottom"` 
(currently the only option), or forced legend to be omitted by using
`legend_position = NA`.

As many of the tick labels would have been very long and would have made the 
plot area are very narrow, we also provided `legend_maxwidth`. This would 
normally cut legend items shorter, but with character strings in y-axis, it 
shortens those too with an ellipsis to a maximum character length we provided. 
Full texts can still be seen on hover.

Next we will create another plot with only a single observation and different 
dimensions, this time a line plot.

```{r producer price index}
data_get("tidy/thi") |>
  filter(
    Indeksisarja == "Teollisuuden tuottajahintaindeksi",
    Tiedot == "Pisteluku (2015=100)",
    Toimiala == "Yhteensä"
  ) |>
  mutate(value = 100 * (value / lag(value, 12) - 1),
         Indeksisarja = "Producer Price Index for Manufactured Products") |>
  drop_na() |>
  roboplot(Indeksisarja,
    "Producer Price Index for Manufactured Products",
    "Annual change, %",
    "Statistics Finland",
    height = 440,
    width = 550
  )
```

We needed to drop values of NA that appeared in data transformation, as
otherwise the plot x-axis would have shown an unnecessary gap on the left side
of the plot. `roboplot()` will omit rows from data where all values are NA, but
keeps them otherwise.

We decided to provide a column name for the parameter `color`, but as the plot
takes the hovertemplate text from data, we had to translate the column text 
manually into English to match the plot. We could have just not have provided 
`color` just as well as there is only a single trace.

We also wanted a plot with different dimensions, and we give the the same 
dimensions that we decided the narrow png images will have, so the output
and png files match.

As you can see from the next few examples, all of this makes the workflow for
createing dynamic plots very short and simple.

```{r fao}
data_get("tidy/fao_food_price_index") |>
  filter(Type == "Nominal",Variable == "Food Price Index") |>
  roboplot(
    Variable,"FAO Monthly Real Food Price Index","Index, 2014–2016=100","FAO"
    )

```

## Controlling x-axis format

The next two plots are weekly and daily data, and `roboplot()` tries to guess
the correct date format from data frequency and zoom level. Sometimes you have
a specific format in mind, and we'll cover these next.

```{r petrol}
data_get("tidy/ec_oil_bulletin") |>
  filter(
    Country %in% c("FI","SE", "DE"), 
    Variable == "Euro-super 95  (I), 1000L", 
    Taxes == T, 
    time >= "2018-01-01") |>
  mutate(
    value = value / 1000,
    Country = recode(Country, FI = "Finland", DE = "Germany", SE = "Sweden")) |>
  roboplot(
    Country,
    "Petrol price, 95E10",
    "Euros per litre",
    "European Commission",
    height = 440,
    width = 550,
    plot_axes = set_axes(xformat = "%Y")
  )

```


```{r electricity}
d <-
  str_c("entsoe/dap_", c("FI","DE_LU","SE3","FR")) |>
  map(data_get, start_time = "2020-01-01") |>
  bind_rows() |>
  arrange(time) |>
  group_by(Area) |>
  mutate(
    ma28 = slider::slide_index_dbl(value, time, mean, .before = days(28))
  ) |>
  group_by(Area, time = floor_date(time, "days") |> as_date()) |>
  summarise(value = mean(ma28), .groups = "drop") |>
  arrange(Area, time)

d |>
  roboplot(Area,"Electricity prices, 28-day moving average","€/MWh","Entso-E")
```

We wanted to to ensure these x-axes are always displayed in years for the first
two plot and month / year for the plot with daily data. Using years only for the
last We provided `xformat` in `plot_axes`. The formatting used is 
[d3 time format](https://github.com/d3/d3-3.x-api-reference/blob/master/Time-Formatting.md). 
You can control y-axis with `yformat` [d3 number format](https://github.com/d3/d3-3.x-api-reference/blob/master/Formatting.md#d3_forma),
or use the appropriate format if you use something other than date data for 
x-axis or numeric data for y-axis. Character formatting is not currenlty an
option.

## Control trace order

`roboplot()` uses the mean of column `value` if it exists to factorize the column
used for `color` if it's not factorized beforehand, and displays the trace and
legend in that order. If you want to specify another order you have to transform
the data itself as in the examples that follow.

```{r sentiment_average}
d <- data_get("ec/esi_nace2") |>
  filter(
    Country %in% 
      c("European Union (current composition)", "Finland", "Germany", "Sweden"),
    Indicator == 
      "The Economic sentiment indicator is a composite measure (average = 100)",
    time >= "2017-01-01"
  ) |>
  mutate(
    Country = 
      recode(Country, "European Union (current composition)" = "EU") |>
      fct_relevel("EU","Finland"),
    Indicator = "Economic sentiment indicator"
    )

d |> roboplot(
    Country,
    "Economic sentiment indicator",
    "Score, 100 = long-run average",
    "European Commission"
  )
```


```{r consumer_confidence_eu}
data_get("ec/esi_nace2") |> 
  filter(Country %in% c("European Union (current composition)", "Finland"),
         Indicator == "Consumer confidence indicator (20%)",
         time >= "1996-01-01"
         ) |> 
  mutate(
    Country = recode(Country, "European Union (current composition)" = "EU") |>
      fct_relevel("EU","Finland")
    ) |>
  roboplot(
    Country, "Consumer confidence indicator","Score","European Commission"
    )
```

Note that `forcats::fct_relevel2()` from forcats behaves differently than `forcats::fct_relevel()`.

```{r consumer_confidence}
d <-
  data_get("oecd/KEI", dl_filter = "CSCICP02.FIN+CHN+USA+EA19+JPN") |>
  filter(FREQUENCY == "Monthly")
attr(d, "frequency") <- "Monthly"

d |>
  select(LOCATION, time, value) |>
  filter(time >= "2000-01-01") |>
  mutate(LOCATION = recode(LOCATION,
    "United States" = "USA",
    "China (People's Republic of)" = "China",
    "Euro area (19 countries)" = "Euro area"
  )) |>
  with_groups(LOCATION, mutate,
    value = 10 * scale(value)[, 1] + 100,
    value = 100 * value / mean(value[year(time) == 2019])
  ) |>
  filter(time >= "2018-01-01") |>
  mutate(LOCATION = fct_reorder2(LOCATION, value, time,.desc = F)) |> 
  roboplot(
    LOCATION,"Consumer Confidence Indicator","Index, 2019 = 100","OECD"
    )
```

## Zeroline control

For the next few plots we want to use a stronger zeroline than what the default 
plotly plots have, so we will use the `zeroline` parameter. Zeroline can also 
take a numeric value which locates the zeroline at the given y-axis location.

```{r, short_term_interest}
data_get("tidy/euribor") |>
  filter(str_detect(Korko, "(1|6|12) kk.*360")) |> 
  mutate(Korko = str_remove(Korko,"\\(.*") |> fct_inorder()) |>
  filter(time >= "2014-01-01") |>
  roboplot(Korko,"Short-term interest rates, Euribor","%","Bank of Finland",
    zeroline = TRUE
  )
```

```{r bond_rates}
data_get("tidy/10yield") |>
  drop_na() |>
  filter(time >= "2012-01-01") |>
  select(Maa, time, value) |>
  mutate(Maa = recode(Maa, "Suomi" = "Finland", "Saksa" = "Germany")) |>
  roboplot(
    Maa,
    "Long-term government bond rates",
    "%, 10-year bond",
    caption = set_caption(template = "Sources: Bank of Finland, Deutche Bundesbank & Fed."),
    zeroline = 2)
```

## More axis controls

For the next plot we need to plot data other than dates as x-axis values.

```{r yield_curves}
data_get("ecb/YC",
     dl_filter = "B.U2.EUR.4F.G_N_A.SV_C_YM.",
     query_args = list(startPeriod = "2022-01-01"),
     labels = FALSE
     ) |>
  filter(`Financial market data type` |> str_starts("SR_")) |>
  filter(
    time %in% c(max(time), min(time), as.Date(c("2022-02-28","2022-06-01")))
  ) |>
  mutate(
    Year = str_match(`Financial market data type`, "([[:digit:]]*)Y")[,2],
    Month = str_match(`Financial market data type`, "([[:digit:]]*)M")[, 2]
  ) |>
  mutate(across(c(Year, Month), as.integer)) |>
  replace_na(list(Year = 0, Month = 0)) |>
  mutate(Maturity = Year + Month / 12) |>
  select(Maturity, time, value) |>
  mutate(Time = as.character(time)) |>
  arrange(Maturity, value) |>
  roboplot(
    Time,
    "Yield curves for AAA euro area government bonds",
    "% interest rate vs. maturity",
    caption = set_caption(
      template = "The curves show the yield curve at various points in time.<br>Source: ECB."
    ),
    plot_axes = set_axes(x = "Maturity", xticktype = "numeric"),
    height = 440,
    width = 550
  )
```

We used `plot_axes` in a similar way we used it with the previous horizontal bar
plot, but this time we defined the data column used for x-axis, and provided
the type `roboplot()` will use for tick formatting. Note that to ensure the 
traces are drawn correctly, you have to order the data yourself before using it
it in `roboplot()`.

```{r unemployment, eval = F, include = F}
data_get("eurostat/une_rt_m") |>
  filter(s_adj == "Seasonally adjusted data, not calendar adjusted data",
         age == "Total", 
         unit == "Percentage of population in the labour force",
         sex == "Total",
         str_detect(geo, "Fin|Europ|United S|Jap|Swe"),
         time >= "2019-01-01") |>
  mutate(
    geo = recode(geo, "European Union - 27 countries (from 2020)" = "EU")) |>
  roboplot(
    geo,"Unemployment rate","%, seasonally adjusted","Eurostat",
    height = 440,
    width = 550
    )

```

For this plot we also wanted to show the numeric y-axis with logarithmic scale.

```{r bitcoin, eval = F, include = F}
#https://data.nasdaq.com/data/BCHAIN/MKTCP-bitcoin-market-capitalization
Quandl::Quandl("BCHAIN/MKTCP") |>
  as_tibble() |>
  rename(time = Date) |>
  rename(value = Value) |> 
  filter(year(time) >= 2017) |>
  mutate(value = value / 1000000000) |>
 # mutate(time = year(time)) |>
  roboplot(
    title = "Bitcoin Market Capitalization",
    subtitle = "Billion dollars",
    caption = "Nasdaq Data Link",
    plot_axes = set_axes(xformat = "%Y", yticktype = "log")
  )
```

```{r china_export}
# China housing investment, from National bureau of statistics of China
# Total Value of Exports, Current Period(1,000 US dollars)
# https://data.stats.gov.cn/english/easyquery.htm?cn=A01
d <- tribble(
  ~time, ~value,
  "Oct 2021", 300221418,
  "Nov 2021", 325525287,
  "Dec 2021", 340498780,
  "Jan 2022", 327285953,
  "Feb 2022", 217416504,
  "Mar 2022", 276084613,
  "Apr 2022", 273619664,
  "May 2022", 308244886,
  "Jun 2022", 331264197,
  "Jul 2022", 332964256,
  "Aug 2022", 314920505, 
  "Sep 2022", 322755333,
  "Oct 2022", 298371747
) |> mutate(value = value / 1000000,
            )


d |> 
  roboplot(
    title = "Value of monthly exports from China", 
    subtitle = "Billion US dollars",
    caption = "National Bureau of Statistics, China", 
    plot_axes = set_axes(xticktype = "character")
    )
```

Finally, for the last plot we had a `time` column which was not in date format
and we did not want to transform it into such, so we defined the ticktype for 
x-axis as `"character"`.

Now we only have to check if the png files are in the directory we wanted them 
to be in.

```{r png_check}

list.files(str_c(tempdir(),"/plots")) |> 
  str_subset("png$") |> 
  str_c(collapse = "<br>") |>
  # roboplotr:::roboplotr_combine_words() |> 
  htmltools::HTML()

```

Everything seems to be in order! Now you have an html-document with all the 
plots needed for your report, and .png files of every plot stored, ready to be
used in another document. You can do a lot more fine-tuning with roboplotr, 
and every function of the package has its own documentation that should give
you a head start.
